Regression example plots (in-class)
==================================================================================
author: Bernhard Angele
date: 22 September 2016


```{r, echo=FALSE}
library(knitr)
n <- 30
intercept <- 65
options(digits = 3, scipen = 4)

set.seed("126")
cat_weight <- rnorm(n, mean = 5, sd = 1) # cat weight in kg
cat_age <- rnorm(n, mean = 60, sd = 24) # cat age in months
cat_breed <- factor(c(rep("Shorthair",n/2), rep("Manx", n/2)))

food_eaten <- intercept + 20*(cat_weight - mean(cat_weight)) + 2 + .5*(cat_age - mean(cat_age)) + rnorm(n, mean = 0, sd = 10)
catfood <- data.frame(CatWeight = cat_weight, FoodEaten = food_eaten)
catfood_age <- data.frame(CatWeight = cat_weight, CatAge = cat_age, FoodEaten = food_eaten)


print_p <- function(p){
  if(p < .01) return("*p* < .01")
  if(p <= .05) return(paste("*p* =", p))
  if(p > .05) return("*p* > .05")
}

Question 3
========================================================


```{r, echo=FALSE}
x <- seq(-3, 3, .1)
y <- 1 + x + rnorm(length(x),0,4)
plot(x, y, axes = FALSE)
grid(col = "lightgray", lty = "dotted",
     lwd = par("lwd"), equilogs = TRUE)
axis(1, pos = 0, at = -5:5)
axis(2, pos = 0, at = seq(from = -10, to = 10, by = 2))
abline(0.5, -1, col = "red")
```


Question 4
========================================================


```{r, echo=FALSE}
x <- seq(-5, 5, .1)
y <- 1 - 0.25*x + rnorm(length(x),0,2)
plot(x, y, axes = FALSE)
grid(col = "lightgray", lty = "dotted",
     lwd = par("lwd"), equilogs = TRUE)
axis(1, pos = 0, at = -5:5)
axis(2, pos = 0, at = seq(from = -20, to = 20, by = 2))
abline(lm(y ~ x), col = "red")
```


Question 5
========================================================


```{r, echo=FALSE, fig.width=10, fig.height=8}
x <- seq(-5, 5, .1)
y <- 1 - 0.25*x + rnorm(length(x),0,2)
plot(x, y, axes = FALSE)
grid(col = "lightgray", lty = "dotted",
     lwd = par("lwd"), equilogs = TRUE)
axis(1, pos = 0, at = seq(-5,5, .5))
axis(2, pos = 0, at = seq(from = -20, to = 20, by = 1))
abline(lm(y ~ x), col = "red")
```


Question 6
========================================================


```{r, echo=FALSE, fig.width=10, fig.height=8}
x <- seq(-5, 5, .1)
y <- 3 + x + rnorm(length(x),0,4)
plot(x, y, axes = FALSE)
grid(col = "lightgray", lty = "dotted",
     lwd = par("lwd"), equilogs = TRUE)
axis(1, pos = 0, at = seq(-5,5, .5))
axis(2, pos = 0, at = seq(from = -20, to = 20, by = 1))
abline(3,1, col = "red")
```

Question 7
========================================================
```{r, echo=FALSE, fig.width=10, fig.height=8}
x <- seq(-5, 5, .1)
y <- 2 + rnorm(length(x),0,4)
plot(x, y, axes = FALSE)
grid(col = "lightgray", lty = "dotted",
     lwd = par("lwd"), equilogs = TRUE)
axis(1, pos = 0, at = seq(-5,5, .5))
axis(2, pos = 0, at = seq(from = -20, to = 20, by = 1))
abline(-.5,0, col = "red")
```

