---
title: "Lecture 7 Seminar slides with answers"
author: "Regression"
date: "Press F for fullscreen"
output: 
  revealjs::revealjs_presentation:
    self_contained: true
    transition: fade
    css: "robot-lung.css"
---

# Understanding contrasts a bit more

```{r prepare, echo = F, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, out.height = "50%", fig.width = 8, fig.height = 4.5, cache = TRUE)
library(report)
library(MASS)
library(jmv)
library(gamlj)
library(haven)
library(tidyverse)
library(sjPlot)
library(equatiomatic)
```

-   For this seminar, we will use an invented data set that will let you experience a bit more about how contrasts work
-   Please download the following data set from Brightspace and open it in jamovi/SPSS: fluffy.csv
    -   For SPSS, you may have to use the CSV import wizard (see [here](https://libguides.library.kent.edu/SPSS/ImportData) for detailed instructions)
    -   jamovi will open the file immediately without issues.
-   These (fictitious) data are from an experiment in which participant rate the positivity of emotions triggered by looking at photos of different dogs and cats (on a scale from 0 - 100, where 0 is most negative and 100 is most positive).
-   An unrelated group of participants rated the same photos in terms of how fluffy they thought the animal was (also on a scale from 0 to 100, where 0 is sandpaper and 100 is extra-soft plush fabric).
-   Also, the file notes whether the photo showed a cat or a dog.

# Your task

-   First, perform a regression analysis where you predict positive feelings only by fluffiness. Note down the results, and write one sentence to summarise what they suggest.

-   Then, perform a regression analysis where, in addition to fluffiness, you also add animal type (cat vs. dog) as a discrete predictor. How is this model different from the first one? How can you explain the difference?

```{r read_data, include=FALSE}
fl <- read_csv("fluffy.csv")
```

# Set up the tests

```{r prepare_data}
fl <- fl %>% mutate(animal = factor(animal, levels = c("cat", "dog")))
#contrasts_for_animal <- cbind(c(0, 1))
#contrasts(asc$Group_factor) <- contrasts_for_group %>% ginv %>% t %>% zapsmall %>% `colnames<-`(c("noSH_vs_SH", "HistoricSH_vs_CurrentSH"))

```

-   Start by fitting a regression model on `positive_feelings` with just `fluffiness`
    -   You can do that in the `Linear Regression` module in jamovi.
        -   In SPSS, use `General Linear Model` → `Univariate...` so it is easier to add the factor later.

# Results: Simple regression

-   Note the slope for `fluffiness`.

```{r simple_regression}
lm1 <- lm(data = fl, positive_feelings ~ fluffiness)
tab_model(lm1, show.stat = TRUE)
fl$predicted1 <- predict(lm1)
```

-   In words: `r report(lm1)`

# Now add the contrast for animal

-   In SPSS/jamovi: add `animal` to `Factors`
-   What happens to the slope for `fluffiness`?

# Question

-   Why was the slope for `fluffiness` negative when only `fluffiness` was in the model, but positive when we have both `fluffiness` and `animal` in the model?
-   How is this related to the idea of "statistically controlling" for a variable?
-   Hint: plotting might help -- think of Anscombe's Quartet!

# Final activity (if there is time)

- Use a contrast of -1 and 1 for `animal` instead of 0 and 1.
- What changes?
- Write the regression equation and plug in the dummy values.
- Why is the coefficient in the -1 and 1 model exactly half of the coefficient in the 0 and 1 model?
