---
title: 'Homework 3 example: The effect of swear word use on pain tolerance'
author: "Bernhard Angele"
date: "16/10/2014"
output: word_document
---

### Results



```{r, echo = FALSE}
# echo = FALSE will keep the code and its output from appearing in the final document.
# don't forget to replace the \ backslashes with forward slashes / if you're on Windows
library(ez)
library(knitr)

# round all numbers to the 2nd digit
options(digits = 2)

#setwd("C:/replace_this_with_your_actual_working_directory/")

pain <- read.csv("pain.csv")

# add a subject column for ezANOVA
# very simple: each observation is a different person
# "subject" is a discrete variable, so make it a factor()
pain$subject <- factor(1:nrow(pain))

pain_stats <- ezStats(data = pain, 
                       dv = Time.In.Cold.Water, 
                       wid = subject, 
                       between =  Swear.Condition)
```

In order to evaluate the effect of swear word usage on pain tolerance, we performed a one-way between-subjects ANOVA with time until hand withdrawal as the dependent variable and swear word usage (quiet vs. neutral word vs. swear word) as the independent variable. Table 1 and Figure 1 show the condition means.

*Table 1. Means, sample sizes (N), standard deviations (SD), and standard errors (SE) for the swear word usage conditions.*
```{r, results='asis', echo=FALSE}
# need results = "asis" so that the table is drawn properly
# get rid of the last column of pain_stats (you don't need Fisher's LSD in a results section)
pain_stats <- pain_stats[,1:4]
# Rename the first column to make the table look more professional
colnames(pain_stats)[1] <- "Condition"
# instead, add a standard error column
pain_stats$SE <- pain_stats$SD/sqrt(pain_stats$N)
# make a nice-looking table  
kable(pain_stats)
```

*Figure 1. Condition means. Error bars denote 95% confidence intervals.*

```{r, echo=FALSE}
# We could use ezPlot to make the figure, but we'd rather have a barplot
# use ggplot2 to plot pain_stats
library(ggplot2)
# calculate upper and lower bounds of 95% CIs around the condition means for error bars
pain_stats$upper <- pain_stats$Mean + 1.96 * pain_stats$SE
pain_stats$lower <- pain_stats$Mean - 1.96 * pain_stats$SE

pain_plot <- qplot(data =  pain_stats, x = Condition, y = Mean, fill = Condition, geom = "bar", stat = "identity", ymax = upper, ymin = lower)
# geom = "bar": draw a barplot
# stat = "identity": don't aggregate the data (they are already aggregated by ezStats)
# fill = colours of the bars
# ymin and ymax: lower and upper limits of the error bars

pain_plot + 
  geom_errorbar(width = .25)+ # add error bars
  ylab("Mean time until hand withdrawal in seconds") # change y axis label to something more informative

```

```{r, echo=FALSE}
# perform the actual ANOVA so we can use the results when writing our report
pain_anova <- ezANOVA(data = pain, 
                      dv = Time.In.Cold.Water,
                      wid = subject, 
                      between = Swear.Condition, 
                      return_aov = TRUE)

# get the residuals for normality check
pain_anova_resid <- resid(pain_anova$aov)

# Perform the Shapiro-Wilk test
pain_anova_shapiro <- shapiro.test(pain_anova_resid)

# do pairwise t-tests
pain_anova_pairwise <- pairwise.t.test(x = pain$Time.In.Cold.Water, g = pain$Swear.Condition, pool.sd = FALSE)

# here's a little helper function to get the p-values printed right
print_p <- function(p){
  if(p < .01) return("*p* < .01")
  if(p <= .05) return(paste("*p* =", p))
  if(p > .05) return("*p* > .05")
}
```

The ANOVA showed that overall, there was a significant effect of the swear condition, *F*(`r pain_anova$ANOVA$DFn`, `r pain_anova$ANOVA$DFd`) = `r pain_anova$ANOVA$F`, $\eta^2$ = `r pain_anova$ANOVA$ges`, `r print_p(pain_anova$ANOVA$p)`. There was no evidence for unequality of variances from Levene's test (`r print_p(pain_anova$"Levene's Test for Homogeneity of Variance"$p)`). Also, the Shapiro-Wilk test revealed no deviations from normality in the residuals (`r print_p(pain_anova_shapiro$p.value)`).

Post-hoc comparisons performed using Holm-Bonferroni corrected *t*-tests showed that the quiet condition (mean time until hand withdrawal: `r pain_stats[pain_stats$Condition == "Quiet",]$Mean` s) and the neutral word condition (mean: `r pain_stats[pain_stats$Condition == "Neutral Word",]$Mean` s) were significantly different from the swear word condition (mean: `r pain_stats[pain_stats$Condition == "Swear Word",]$Mean` s; quiet vs. swear word: `r print_p(pain_anova_pairwise$p.value["Swear Word", "Quiet"])`; neutral vs. swear word: `r print_p(pain_anova_pairwise$p.value["Swear Word", "Neutral Word"])`), but the neutral word condition did not differ significantly from the quiet condition (`r print_p(pain_anova_pairwise$p.value["Quiet","Neutral Word"])`).

### Summary

Based on the results of our analyses, we can conclude that repeating swear words seemed to increase participants' pain tolerance compared to repeating neutral words or remaining quiet. Repeating neutral words did not seem to increase participants' pain tolerance compared to remaining quiet. It seems that using swear words in painful situations may be an effective strategy.