---
title: "Advanced Statistics Practice Test"
author: "Bernhard Angele"
subtitle: Model solution
output: pdf_document
---

# Part 1 -- Data analysis (50 %)

## Scenario
The university wants to find out whether therapy animals are useful in lowering student stress ahead of exams. In order to test this, they asked 24 students to each have three different exam revision sessions: one with no therapy animals present, one with a therapy cat, and one with a therapy dog. The students were then asked to report their stress levels on a scale from 0 to 100, where 0 is no stress at all and 100 is extreme stress.

## Assignment
Conduct the appropriate ANOVA and the corresponding post-hoc *t*-tests comparing all three factor levels to each other (assume all assumptions are met). In order to help you with this, a number of intermediate values have already been computed for you below. Report the results of the ANOVA and *t*-tests as one would in the results section of an academic manuscript (using APA style). Finally, in layperson (non-academic) language describe the results and summarise what you can conclude about the effect of therapy dogs and cats. There is no word limit, but the whole report should be rather brief. You may use a calculator, Microsoft Excel, or any other computer programme, as well as your pet notes and any other offline resources you wish to bring to the test. Your test must be your own work and you may not collaborate with anyone else. Show your work so you can get partial credit in case you make small calculation mistakes.

## Data

In order to save you time, part of the ANOVA table is provided below, as well as the group means and standard deviations and the means and standard deviations of the differences between factor levels for each student.

```{r, echo=FALSE, results='asis', warning=FALSE}
rm(list = ls())
library(knitr)
df <- read.csv("Practice_assignment_data.csv")[,-1]

SS_tot = sum((df$stress - mean(df$stress)) ^ 2)

subject_means <- tapply(df$stress, df$subject, mean)

pet_means <- tapply(df$stress, df$pet, mean)
pet_sd <- tapply(df$stress, df$pet, sd)
pet_n <- tapply(df$stress, df$pet, length)

pet_table <- data.frame(pet = names(pet_means), Mean = pet_means, SD = pet_sd, N = pet_n)

calc_differences <- function(Condition1 = "Dog", Condition2 = "Cat", data = df, digits = 3){
  differences <- subset(df, pet == as.character(Condition1))$stress - subset(df, pet == as.character(Condition2))$stress
  data.frame(m_diff = signif(mean(differences), digits), sd_diff = signif(sd(differences), digits))
}

differences_table <- data.frame(L1 = c("Cat", "Dog", "Cat"), L2 = c("No animal", "No animal", "Dog"))

differences_table <- cbind(differences_table,t(with(differences_table, mapply(calc_differences, L1, L2))))

SS_S = length(unique(df$pet)) * sum((subject_means - mean(df$stress))^2)

SS_A = length(unique(df$subject)) * sum((pet_means - mean(df$stress))^2)

SS_AxS <- SS_tot - SS_A - SS_S

SS_WS <- SS_A+SS_AxS

df_tot <- length(unique(df$pet)) * length(unique(df$subject)) - 1

df_A <- length(unique(df$pet)) - 1

df_S <- length(unique(df$subject)) - 1

df_AxS <- df_A * df_S

df_WS <- df_A + df_AxS

MS_A <- SS_A/df_A
MS_S <- SS_S/df_S
MS_AxS <- SS_AxS/df_AxS
MS_WS <- SS_WS/df_WS

F_A <- MS_A/MS_AxS
options(digits = 3)

library(ez)
df.aov <- ezANOVA(data = df, dv = stress, wid = subject, within = (pet), return_aov = TRUE)

print_p <- function(p){
  if(p < .01) return("*p* < .01")
  if(p <= .05) return(paste("*p* =", round(p)))
  if(p > .05) return("*p* > .05")
}
```

### Means table
```{r, echo=FALSE}

 kable(pet_table, row.names = F, col.names = c("Condition", "Mean", "SD", "N"))
```

### Differences between groups
```{r, echo = FALSE}
kable(differences_table, col.names = c("$Condition_1$", "$Condition_2$", "$Mean(Condition_1 - Condition_2)$", "$SD(Condition_1 - Condition_2)$"), align = c('l','l','r','r'))

```

### ANOVA Table

|Source            |SS            |df              |MS        |*F*    |*p* |
|:-----------------|:-------------|:---------------|:---------|:------|:---|
|Between Subjects  |`r SS_S`      |`r df_S`        |`r MS_S`  |       |    |
|Within Subjects   |`r SS_WS`     |`r df_WS`       |`r MS_WS` |       |    |
|--Animal condition|`r SS_A`      |`r df_A`        |`r MS_A`  |`r F_A`|`r print_p(pf(q = F_A, df1 = df_A, df2 = df_AxS, lower.tail = FALSE))`|
|--Residual        |`r SS_AxS`    |`r df_AxS`      |`r MS_AxS`|       |    |
|**Total**         |**`r SS_tot`**|**`r df_tot`**  |          |       |    |


### Paired t-tests (post-hoc)
```{r, echo = FALSE}
t_cat_dog <- t.test(subset(df, pet == "Cat")$stress, subset(df, pet == "Dog")$stress, paired = TRUE)
t_dog_noanimal <- t.test(subset(df, pet == "Dog")$stress, subset(df, pet == "No animal")$stress, paired = TRUE)
t_cat_noanimal <- t.test(subset(df, pet == "Cat")$stress, subset(df, pet == "No animal")$stress, paired = TRUE)

print_p <- function(p){
  if(p < .01) return("*p* < .01")
  if(p <= .05) return(paste("*p* =", round(p, 3)))
  if(p > .05) return("*p* > .05")
}

n <- length(unique(df$subject))
```

Remember, for the paired t-test, we calculate the differences between the means for each participant and use a t-test to determine whether we can reject the $H_0$ that the true population mean of these differences is 0 ($H_0: \mu_{d0} = 0$)

If the $H_0$ is true, the t-value is calculated as follows: $t_{n-1} = \frac{\bar{d} - \mu_{d0}}{\hat{\sigma}_{\bar{d}}} = \frac{\bar{d} - 0}{\hat{\sigma}_{\bar{d}}} = \frac{\bar{d}}{\hat{\sigma}_{\bar{d}}}$

We can estimate the standard error of the difference mean $\hat{\sigma}_{\bar{d}}$ from the standard deviation of the difference: $$\hat{\sigma}_{\bar{d}} = \frac{s_d}{\sqrt{n}}$$

Plugging this into the equation for *t*, we get: $$t_{n-1} =  \frac{\bar{d}}{\frac{s_d}{\sqrt{n}}}$$

$t_{Dog-Cat} = \frac{\bar{d}}{\frac{s_{d}}{\sqrt{n}}} = \frac{`r subset(differences_table, L1 == "Cat" & L2 == "Dog")$m_diff`}{\frac{`r subset(differences_table, L1 == "Cat" & L2 == "Dog")$sd_diff`}{\sqrt{`r n`}}} = `r (t_dc <- subset(differences_table, L1 == "Cat" & L2 == "Dog")$m_diff[[1]] / (subset(differences_table, L1 == "Cat" & L2 == "Dog")$sd_diff[[1]]/sqrt(n)))`$

$t_{Dog-No animal} = \frac{\bar{d}}{\frac{s_{d}}{\sqrt{n}}} = \frac{`r subset(differences_table, L1 == "Dog" & L2 == "No animal")$m_diff`}{\frac{`r subset(differences_table, L1 == "Dog" & L2 == "No animal")$sd_diff`}{\sqrt{`r n`}}} = `r (t_dn <- subset(differences_table, L1 == "Dog" & L2 == "No animal")$m_diff[[1]] / (subset(differences_table, L1 == "Dog" & L2 == "No animal")$sd_diff[[1]]/sqrt(n)))`$

$t_{Cat-No animal} = \frac{\bar{d}}{\frac{s_{d}}{\sqrt{n}}} = \frac{`r subset(differences_table, L1 == "Cat" & L2 == "No animal")$m_diff`}{\frac{`r subset(differences_table, L1 == "Cat" & L2 == "No animal")$sd_diff`}{\sqrt{`r n`}}} = `r (t_cn <- subset(differences_table, L1 == "Cat" & L2 == "No animal")$m_diff[[1]] / (subset(differences_table, L1 == "Cat" & L2 == "No animal")$sd_diff[[1]]/sqrt(n)))`$

The df for each of these t-values is $`r df_S`$. When determining *p*-values, you must correct for multiple comparisons by either multiplying the observed *p*-value by the number of comparisons or dividing the critical *p*-value ($\alpha = .05$) by the number of comparisons (in this case, 3).

### Model writeup
In order to evaluate the effect of therapy animals on stress, we performed a one-way within-subjects ANOVA with stress rating (on a scale from 0 to 100) as the dependent variable and therapy animal (no animal vs. cat vs. dog) as the within-subjects independent variable.
The ANOVA showed that overall, there was a significant main effect of therapy animal, *F*(`r df_A`, `r df_AxS`) = `r F_A`, `r print_p(pf(F_A, df_A, df_AxS, lower.tail = FALSE))`. Post-hoc paired *t*-tests showed that participants reported higher stress after a revision session with no animal than after one with a cat, *t*(`r t_cat_noanimal$parameter`) = `r t_cn`, `r print_p(t_cat_noanimal$p.value * 3)`, and after a session with a dog *t*(`r t_dog_noanimal$parameter`) = `r t_dn`, `r print_p(t_dog_noanimal$p.value * 3)`. There was no significant difference between the amount of stress reported after a session with a cat and a session with a dog, *t*(`r t_cat_dog$parameter`) = `r t_dc`, `r print_p(t_cat_dog$p.value * 3)` (all *p*-values reported are Bonferroni-corrected). In conclusion, it seems that both having a therapy cat and having a therapy dog present will reduce stress levels in students compared to having no animal present, while there was no difference between having a dog present and having a cat present. The choice of therapy cat or therapy dog (Note: I'll give credit for any conclusion here, as long as it doesn't contradict the pattern in the ANOVA and the *t*-tests).
